---
title: "Assignment 05"
author: "Shubham Maheshwari - 8894405"
date: "2024-12-01"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

cat("\014") 
if (!is.null(dev.list())) dev.off()
rm(list = ls())
options(scipen = 9)

#If the library is not already downloaded, download it

if(!require(lattice)){install.packages("lattice")}
library("lattice")

if(!require(pastecs)){install.packages("pastecs")}
library("pastecs")

if(!require(corrgram)){install.packages("corrgram")}
library("corrgram")

if(!require(tinytex)){install.packages("tinytex")}
library("tinytex")

if(!require(ggplot2)){install.packages("ggplot2")}
library("ggplot2")

if(!require(polycor)){install.packages("polycor")}
library("polycor")

if (!require(klaR)) {install.packages("klaR")}
library(klaR)

if(!require(partykit)){install.packages("partykit")}
library(partykit)
```

# 1. Preliminary Data Preparation
```{r}
setwd('C:/PROG8435 Data Analytics/Assignment 5')

df <- read.table("PROG8435-24F-Assign05.txt", sep = ",", header = TRUE)

df <- as.data.frame(df)
head(df)

# Rename All Variables to SM 
colnames(df) <- paste(colnames(df), "SM", sep = "_")
head(df)

df <- as.data.frame(unclass(df), stringsAsFactors=TRUE)
head(df)
```

```{r}
# Creating a new variable PC_SM 
df$PC_SM <- as.factor(ifelse(df$Prc_SM < 2251,"L","H"))
head(df)


# Removing PRC Variable 
df <- df[,-c(1)]
head(df)
```
```{r}
summary(df)
table(df$PC_gp)
```

# 2. Exploratory Analysis

```{r}
# Boxplots 
par(mfrow=c(2,2))
for (i in 1:ncol(df)) {
  if (is.numeric(df[,i])) {
    boxplot(df[i], main=names(df)[i],
            horizontal=TRUE, pch=10)
  }
}
par(mfrow=c(1,1))

# Barplots for Factor Columns
par(mfrow=c(2,2))

for(i in 1:ncol(df)) {
  if(is.factor(df[,i])){
    ct <- table(df[i])
    barplot(ct, main=names(df[,i]))
  }
}

par(mfrow=c(1,1))
```
Observation: 
There are clear outliers in several numeric variables such as (floor_SM, Sqft_SM, Dist_SM). These might need further investigation to determine their validity apart from that the categorical data seems fairly balanced overall, except for the notable difference in the H and L categories.

```{r}
hetcor(df$Bed_SM, df$PC_SM)

hetcor(df$floor_SM, df$PC_SM)

hetcor(df$TotFloor_SM, df$PC_SM)

hetcor(df$Bath_SM, df$PC_SM)

hetcor(df$Sqft_SM, df$PC_SM)

hetcor(df$City_SM, df$PC_SM)

hetcor(df$Comp_SM, df$PC_SM)

hetcor(df$Dist_SM, df$PC_SM)

```
Distance and Price have strong positive correlation (0.2783)
while Bath and Price have strong negative correlation ( -0.3601)

```{r}
# Choosing the sampling rate
sr <- 0.75


n.row <- nrow(df)

set.seed(4405)

training.rows <- sample(1:n.row, sr*n.row, replace=FALSE)

train <- subset(df[training.rows,])

test <- subset(df[-c(training.rows),])

summary(train)
summary(test)
```

# 3. Model Development

### 1. Full Model
```{r}
glm.full <- glm(PC_SM ~ ., family = "binomial", data = train)

# Summarize the model
summary(glm.full)

# Plot Cook's distance for influential points
plot(glm.full, which = 4, id.n = 6, main = "Cook's Distance: Full Model")

```
Observations:

Fisher iterations - 5

AIC - 779.93
 
Residual Deviance - 759.93
 
z-values - 7/10 variables are significant however it fails at the intercept 
 
Parameter Co-Efficients - Some variables have positive correlation such as Bedroom and Distance , while others are negatively correlated. 

There are some influential points but nothing of great significance as they are way under the 0.5 mark. 

### 2. Backward Model 
```{r}
# Perform backward selection
glm.back <- step(glm.full, direction = "backward", trace = TRUE)

# Display the summary of the final model
summary(glm.back)

# Plot Cook's distance for the backward selection model
plot(glm.back, which = 4, id.n = 6, main = "Cook's Distance: Backward Model")


```
Observations: 

Fisher iterations - 5

AIC - 778.82
 
Residual Deviance - 762.82
 
z-values - 6/8 variables passes the hypothesis 
 
Parameter Co-Efficients - Bed and Distance are positively correlated, while other variables are not. We could see city Terranova positively correlated as well which will require further investigation. 

### 4. Based on your preceding analysis, recommend which model should be selected and explain why.

From the above analysis we can conclude that the Backward Model is better than the Full Model 

Both the models have no issues fitting the models, as showcased by fischer of 5 

Back model has Lower AIC - 779.82 vs 779.93 

Residual is slightly higher than that of full model, suggesting minimal loss and also due to having less variables which is explanatory 

Removal of non significant values such as totfloor where the full model gave a positive correlation which was not reciprocated by the floor, indicating errors in the model. 


# Part- B 

## 1. Logistic Regression - Stepwise 

```{r}
start_time <- Sys.time()
step.model <- step(glm.full, direction = "both", trace = TRUE)

end_time <- Sys.time()

Time <- end_time - start_time

summary(step.model)

Time
```
```{r}
# Predict probabilities for training data
response_tr <- predict(step.model, newdata = train, type = "response")

# Classify predictions based on the threshold of 0.5
class_tr <- ifelse(response_tr > 0.5, "H", "L")

# Create confusion matrix for training data
conf_tr <- table(train$PC_SM, class_tr,
                           dnn = list("Actual", "Predicted"))
conf_tr

# Calculate accuracy for training data
accuracy_tr <- (conf_tr[2, 2] + conf_tr[1, 1]) / sum(conf_tr)

# Display rounded accuracy
round(accuracy_tr, 3)

```

```{r}

# Predict probabilities for testing data
response_te <- predict(step.model, newdata = test, type = "response")

# Classify predictions based on the threshold of 0.5
class_te <- ifelse(response_te > 0.5, "H", "L")

# Create confusion matrix for testing data
conf_te <- table(test$PC_SM, class_te,
                 dnn = list("Actual", "Predicted"))
conf_te

# Calculate accuracy for testing data
accuracy_te <- (conf_te[2, 2] + conf_te[1, 1]) / sum(conf_te)

# Display rounded accuracy
round(accuracy_te, 3)

```

## 2. Na誰ve-Bayes Classification
```{r}
# Recording the start time
start_tm <- Sys.time()

# Fit the Naive Bayes model
nb.mod <- NaiveBayes(PC_SM ~ ., data = train, na.action = na.omit)

# Record end time
end_tm <- Sys.time()

# Calculate time taken for model fitting
time_taken <- end_tm - start_tm
time_taken

```


```{r warning=FALSE}
# Train Data
pred_NB <- predict(nb.mod, newdata=train)

Conf_NB <- table(Actual=train$PC_SM , Predicted=pred_NB$class)
Conf_NB

Accuracy <- (Conf_NB[1,1] + Conf_NB[2,2])/sum(Conf_NB)

round(Accuracy,4)

```

```{r warning=FALSE}
# Test Data

pred_NB <- predict(nb.mod, newdata=test)

Conf_NB <- table(Actual=test$PC_SM , Predicted=pred_NB$class)
Conf_NB

Accuracy <- (Conf_NB[1,1] + Conf_NB[2,2]/sum(Conf_NB))

round(Accuracy,4)

```
## 3. Recursive Partitioning Analysis
```{r}
start_time <- Sys.time()

RP.mod <- ctree(PC_SM ~ ., data= train)

end_time <- Sys.time()
  
Time <- end_time - start_time
Time

plot(RP.mod, gp=gpar(fontsize=8))
RP.mod

```


```{r}
# Predict on training data
pred.RP <- predict(RP.mod, newdata= train)

# Confusion matrix
Conf_RP <- table(Actual = train$PC_SM, Predicted = pred.RP)

# Print confusion matrix
Conf_RP

# Calculate accuracy for training data 
Accuracy <- (Conf_RP[1,1] + Conf_RP[2,2]) / sum(Conf_RP)
round(Accuracy, 4)


```

```{r}
# Predict on testing data
pred.RP <- predict(RP.mod, newdata= test)

# Confusion matrix
Conf_RP <- table(Actual = test$PC_SM, Predicted = pred.RP)

# Print confusion matrix
Conf_RP

# Calculate accuracy for testing data 
Accuracy <- (Conf_RP[1,1] + Conf_RP[2,2]) / sum(Conf_RP)
round(Accuracy, 4)

```

## 4.Neural Network Fitting
```{r}
# Install and load the nnet package
if (!require("nnet")) {
  install.packages("nnet")
}
library("nnet")

# Record start time
start_tm <- Sys.time()

# Fit the neural network model
nn.mod <- nnet(
  PC_SM ~ .,          # Response variable and predictors
  data = train,       # Dataset
  size = 4,           # Number of nodes in the hidden layer
  rang = 0.0001,      # Initial random weights range
  maxit = 1200,       # Maximum number of iterations
  trace = FALSE       # Suppress detailed trace output
)

# Record end time
end_tm <- Sys.time()

# Calculate time taken for model fitting
time <- end_tm - start_tm
time


```

```{r}
 nn.mod
 summary(nn.mod)
 nn.mod$wts
```



```{r}
# Predict on training data using the neural network mode
pred.nn <- predict(nn.mod, newdata = train, type = "class")

# Confusion matrix
Conf_NN <- table(Actual = train$PC_SM, Predicted = pred.nn)

# Print confusion matrix
Conf_NN

# Calculate accuracy
Accuracy <- (Conf_NN[1,1] + Conf_NN[2,2]) / sum(Conf_NN)
round(Accuracy, 4)
```

```{r}
# Predict on testing data using the neural network mode
pred.nn <- predict(nn.mod, newdata = test, type = "class")


Conf_NN <- table(Actual = test$PC_SM, Predicted = pred.nn)

#Print confusion matrix
Conf_NN


Accuracy<- (Conf_NN[1,1] + Conf_NN[2,2]) / sum(Conf_NN)
round(Accuracy, 4)
```


## 5. Compare All Classifiers

### 1.Which classifier is most accurate?

Logistic Regression :  25.1%
Naive Bayes : 79.44%
Recursive Partition: 69.58%
Neural Network: 63.5%

Na誰ve Bayes is the most accurate classifier with a testing accuracy of 79.44%, outperforming all other models.

### 2.Which classifier seems most consistent (think train and test)?

We measured consistency by taking the difference between the training and testing accuracy, these were the results: 

Logistic Regression : ~ 1%
Naive Bayes : ~ 4.2%
Recursive Partition: ~ 6.7%
Neural Network: ~ 5.03% 

Na誰ve Bayes is the most consistent classifier, with only a 4.2% difference between training and testing accuracy. 

### 3.Which classifier is most suitable when processing speed is most important?

Here are the results of different models

Logistic Regression : 0.2357 seconds
Naive Bayes: 0.0472 seconds
Recursive Partioning: 0.1705 seconds
Neural Network: 0.1065 seconds 

Na誰ve Bayes is the most suitable when processing speed, as it trained in just 0.0472 seconds.  

### 4.Which classifier minimizes false positives?

Here are the results of different models

Logistic Regression : 112
Naive Bayes : 34
Recursive Partioning:  44
Neural Network: 17

Neural Network minimizes false positives, with only 17 false positives in the testing set


### 5.In your opinion, which classifier is best overall? Make sure you state why.

In my opinion Naiive Bayes is the best overall classifier for the following reasons:  

Highest Accuracy: 79.44%.
Most Consistent: Smallest train-test gap (4.2%).
Fastest Training: 0.0472 seconds, suitable for scenarios requiring quick results.
Balanced Performance: While it does not minimize false positives, its overall performance (accuracy and consistency) outweighs this limitation.

# References 

All the code was taken from the course material provided by the professor. 
Would like to mention Tony and Shubham for their help in clearing my doubts.  

