---
title: "Assignment 2 - Data Transformation, Statistical Inference and Comparisons"
author: "Shubham Maheshwari"
date: "2024-10-05"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=5, fig.height=5,
                      fig.path='Figs/', echo = TRUE)

knitr::opts_knit$set(root.dir = 'C:/PROG8435 Data Analytics/Assignment 2')
```

```{r}
#if(!is.null(dev.list())) dev.off()
#cat("\014") 
#rm(list=ls())
options(scipen=9)
```

```{r}

if(!require(tinytex)){install.packages("tinytex")}
library("tinytex")

if(!require(pastecs)){install.packages("pastecs")}
library("pastecs")

if(!require(lattice)){install.packages("lattice")}
library("lattice")
```

# 1. Data Transformation and Preperation

## 1. Initial Transformation
### a. Rename all variables with your initials appended
```{r}
getwd()
df <- read.table("PROG8435-24F-Assign02.txt", 
                         sep = ",", header = TRUE)
head(df)

colnames(df) <- paste(colnames(df),"SM",sep = "_")
head(df)

str(df)

```

### b. Transform character variables to factor variables.
```{r}
df$Ren_SM<- as.factor(df$Ren_SM)
str(df)
```
## 2. Outliers
```{r}
# We will be checking for any outliers using the boxplot technique and further investigate into it accordingly 

# Running a for loop to generate box plots for numerical values only. 
for (i in 1:ncol(df))
{
  if(is.numeric(df[,i]))
  {
    boxplot(df[i], main = names(df)[i],
            horizontal=TRUE, pch=10)
  }
}

```
Observations from the box plots: 

1. DT: The date ranges from 0 - 60 and detects no anomolity in it 

2. TM ( Time of Day): It shows values that are negative which are anamolous since time cannot be negative, will further 
                      look into it using the density plot. 
                      
3. Temperatues from sensor 1 looks to be in the normal range which can also be seen by the other sensors as well. 

4. No anamolities detected in the fan speeds either. 

```{r}
# Using density plot to check for values that are below 0 

densityplot(df$TM_SM, main = "Time of Day (hour)", pch=6)

# Checking other variables for fun 
densityplot(df$S1_T_SM, pch=6)
densityplot(df$S2_T_SM, pch=6)
densityplot(df$S3_T_SM, pch=6)
densityplot(df$FN_SM, pch=6)
```
Observation:

It confirms that there are 3 negative values, I will be proceeding to delete them as there can't be no negative values for hours. 

```{r}
densityplot( ~ df$TM_SM, pch=6)
nr <- which(df$TM_SM < 0)  #Finds values less than 0 
df <- df[-c(nr),]
densityplot( ~ df$TM_SM, pch=6)

```

## 3. Reduce Dimensionality 

### a. Drop any variables that do not contribute any useful analytical information at all.

```{r}
# Index column doesn't provide us with any usefule analytical information at all so we will be removing it. 

df <- df[-c(1)]
```
### b. Apply the Missing Value Filter to remove appropriate columns of data.
```{r}
summary(df)
```
There are no missing values present in the data. 

### c. Apply the Low Variance Filter to remove appropriate columns of data
```{r}
stat.desc(df)
summary(df)
```
All the columns consists of valuable information so there is no need to drop any columns. 


### d. Apply the High Correlation Filter to remove appropriate columns of data.
```{r}
numeric_cols <- df[,3:11]

cor(numeric_cols,method="spearman")
```
From the data above it's clear that the values for light in Sensor 1 and Sensor 2 are highly correlated, so we will be removing one of them. 
```{r}
df <- df[-c(6)]
head(df)
```
### e. Based on our discussions in class, what are some specific benefits of reducing the dimensionality of this particular dataset? 

In the current data set I had removed 2 variables out of 12 making the entire memory usgage 16.66% less and hence reducing the response time by the 16.66% and bringing improvement in the computational efficiency. 


# Organizing Data 

## 1. Scatter Plots 

### a. Create a histogram for Temperature from Sensor 1.

```{r}
hist(df$S1_T_SM,
     col=2,
     xlab="Temperature Sensor 1 (°C)",
     main="Histogram of Temperature Sensor 1")
```
```{r}
hist(df$FN_SM,
     col=3,  
     xlab="Fan Speed",
     main="Histogram of Fan Speed")
```

```{r}
plot(FN_SM ~ S1_T_SM,
     data = df,
     main = "Temperature Sensor 1 vs Fan Speed",
     xlab = "Temperature Sensor 1 (°C)",
     ylab = "Fan Speed")

# Add a horizontal line at the mean fan speed
abline(h = mean(df$FN_SM), col = "red", lty = 2)
```
### d. What conclusions, if any, can you draw from the chart?
The data from the scatter plot shows to be densely populated in the center between 20 - 25 C and fan speed of 1050 - 1150 

It shows No Linear Relationship between the Temperature and the Fan Speeds as there are many points which shows the opposite. Low fan speed and yet a high temperature or vice verca 


```{r}
print("Pearson Correlation")
round(cor(df$FN_SM, df$S1_T_SM),3)
```
I went with Pearson Method as it measures the linear relationship between two continuous variables

Since r is close to 0, then there is no linear relationship between the variables.

REFERENCE: https://datascientest.com/en/pearson-and-spearman-correlations-a-guide-to-understanding-and-applying-correlation-methods

# 3. INFERENCE

## 1. Normality

### a. Create a QQ Normal plot of for Fan Speed.
```{r}
# QQ plot for Fan Speed
qqnorm(df$FN, main = "Plot of Fan Speed")
qqline(df$FN)
```
### b.Conduct a statistical test for normality on Fan Speed.

```{r}
shapiro.test(df$FN_SM)
```
### c. Is Fan Speed normally distributed? What led you to this conclusion?
QQ Normal shows that the majority of the points lie on the lie but still can not confirm yet however we can say that approximately the Fan Speed in Normally Distributed 

Performing the Shapiro Wilks Test, confirms further, since P value is less than 5% it confirms our suspicion that the Fan speed is not normally distributed 


## 2. Statistically Significant Differences

### a. Compare Temperature from Sensor 2 between ‘New’ and ‘Old’ rooms in your dataset using a suitable hypothesis test.

```{r}
# Performing Normality Tests 

qqnorm(df$S2_T_SM)
qqline(df$S2_T_SM)

shapiro.test(df$S2_T_SM)
```

```{r}
# Performing the T-Test since it matches the assumptions 

t.test(S2_T_SM ~ Ren_SM, data = df, var.equal = TRUE)
```
### b. Explain why you chose the test you did.
I chose to go with t-test for couple of reasons:
1. We have a continuous data so had to compare the means of it . 

2. Since two groups are being compared and the data is normally distributed which meets all the assumptions for T-Test to be performed. 



### c. Do you have strong evidence that Temperature from Sensor 2 is different between new and old rooms?

Yes, the p value is significantly less than 5% that clearly indicates the difference of temperature between the two rooms. 





## 3. Multiple Statistical Differences

### a. Determine if Temperature from Sensor 1 varies by Room Number using ANOVA (statistical) and a sequence of boxplots (graphical).
```{r}
summary(aov(S1_T_SM ~ Room_SM, data=df))
print(" ")

boxplot(S1_T_SM ~ Room_SM, data=df,
        main="Temperature from Sensor 1 by Room",
        xlab = "Room", 
        ylab = "Temperature (°C)",
        range=0)

```
Conclusion: 
A Significant p-value less than 0.05 suggests that temperatures varies across the different rooms. 
Boxplot further confirms visually the difference in temperatures across the rooms. 




### b. Determine if Temperature from Sensor 3 varies by Room Number using ANOVA (statistical) and a sequence of boxplots (graphical).
```{r}
summary(aov(S3_T_SM ~ Room_SM, data=df))
print(" ")

boxplot(S3_T_SM ~ Room_SM, data=df,
        main="Temperature from Sensor 3 by Room",
        xlab = "Room", 
        ylab = "Temperature (°C)",
        range=0)
```
Conclusion: 
A Significant p-value= 0.977 is greater than 0.05 suggests that temperatures are similar across the different rooms. 
Box plot further confirms visually the similarity in temperatures across the rooms. 


